\large{\bf Looking forwards and backwards through locally regulated populations }

### 2023-01-10 forwards-and-backwards.tex
factor of 2
.

    First we prove \eqref{eqn:first_moment_rho}.
    Recall that $\smooth{r} \eta(x) = \int p_{\epsilon^2_r}(x - w) \eta(dw)$,
    where $p_t$ is the density of a Gaussian with mean 0 and variance $t$,
    so that applying Fubini, \eqref{eqn:first_moment_rho} is
    \begin{align*}
        \left|
        \int_{\IR^d}
        \theta
            \int_{\IR^d}
                ( p_{\epsilon_r^2}(y - w) - p_{\epsilon_r^2}(x - w) )
            q_\theta(x, dy)
        \eta(dw)
        \right| .
    \end{align*}
    Write $p_{s, x}(\cdot)$ for the density of a Gaussian
    with mean $s \meanq(x)$ and covariance $\epsilon_r^2 I + s \covq(x)$,
    so that $\int p_{\epsilon_r^2}(y - w) q_\theta(x, dy) = p_{1/\theta, x}(w-x)$.
    It therefore suffices to show that for all $x$ and $w \in \IR^d$,
    there exists $K$ such that
    \begin{align} \label{eqn:goal1}
        \left|
        \theta
            \int_{\IR^d}
                ( p_{\epsilon_r^2}(y-w) - p_{\epsilon_r^2}(x-w) )
            q_\theta(x, dy)
        \right|
        &=
        \theta \left|
            p_{1/\theta, x}(w-x) - p_{0, x}(w-x)
        \right|
        \le
        K p_{\epsilon^2_\gamma}(w-x) .
    \end{align}
    However,
    $\theta( p_{1/\theta, x}(z) - p_{0, x}(z) )
    = \partial_s p_{s, x}(z)$ for some $0 \le s \le 1/\theta$.
    Write $\Gamma(s,x) = \epsilon_r^2 I + s \covq(x)$,
    so that
    \begin{align*}
        p_{s, x}(z)
        &=
        \frac{1}{\left(2 \pi |\Gamma(s,x)|\right)^{d/2}}
        \exp\left(
            -\frac{1}{2} (z - s\meanq(x))^T \Gamma(s,x)^{-1} (z - s\meanq(x))
        \right) ,
    \end{align*}
    and note that if $\lambda_i$ are the eigenvalues of $\covq(x)$
    then $|\Gamma(s, x)| = \prod_i (\epsilon_r^2 + s \lambda_i)$,
    and
    $\partial_s |\Gamma(s, x)| = \sum_i \lambda_i |\Gamma(s, x)| / (\epsilon_r^2 + s \lambda_i)$.
    Therefore,
    \begin{align*}
        \partial_s p_{s, x}(z)
        &=
        \left(
            \meanq(x)^T \Gamma(s,x)^{-1} (z - s \meanq)
            +
            (z - s \meanq(x))^T
            \Gamma(s,x)^{-1} \covq(x) \Gamma(s,x)^{-1}
            (z - s \meanq(x))
        \right. \\ & \qquad \qquad \qquad \left. {}
            -
            \sum_i \frac{\lambda_i}{\epsilon_r^2 + s \lambda_i}
        \right)
        p_{s, x}(z) ,
    \end{align*}
    where $z^T$ is the transpose of $z$.
    This implies that
    \begin{align*}
        \frac{
        \theta
            \int_{\IR^d}
                ( p_{\epsilon_r^2}(y-w) - p_{\epsilon_r^2}(x-w) )
            q_\theta(x, dy)
        }{
            p_{\epsilon^2_\gamma}(w-x) .
        }
        =
        h(z) e^{k(z)},
    \end{align*}
    where $h(z)$ and $k(z)$ are quadratic polynomials in $z$
    whose coefficients depend on $s$ and $x$ but are uniformly bounded for $s \ge 1$
    and $x \in \IR^d$.
    Therefore, \eqref{eqn:goal1} holds for some finite $K$
    if and only if $k(z) < 0$ for all $z$ outside some bounded region.
    This is
    \begin{align*}
        k(z)
        =
        \frac{1}{2\epsilon_\gamma^2} \|z\|^2
        -\frac{1}{2} (z - s\meanq(x))^T \Gamma(s,x)^{-1} (z - s\meanq(x)) ,
    \end{align*}
    which is negative for all $z$ outside a bounded region.
    Since $\inf_z z^T \Gamma(s,x)^{-1} z / \|z\|^2 = 1 / (s \lambda(x) + \epsilon_r^2)$,
    where $\lambda(x) = \sup z^T \covq(x) z / \|z\|^2$ is the largest eigenvalue of $\covq(x)$,
    equation~\eqref{eqn:first_moment_rho} follows from the assumption
    that $\epsilon_r^2 + 2 \sup_x \lambda(x)/\theta < \epsilon_\gamma^2$.
    (Note that we do not yet need the factor of 2.)

@@

\begin{proof}[of Lemma~\ref{lem:tightness_for_F}]
    Let $\bar f(m) = \sup_x \sup_{k \le m} f(x, k)$, which is finite by assumption.
    Since $p_{\sigma^2}*\eta(x) \le \langle 1, \eta \rangle / (2 \pi \sigma^2)^{d/2}$,
    we have that
    $$
        \sup_{0 \le s \le T} \sup_{x \in \IR^d} f(x, p_{\sigma^2}*\eta^N_s(x))
        \le
        \bar f( \sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle  / (2 \pi \sigma^2)^{d/2} ) .
    $$
    This implies that if $0 \le s \le T$ then
    $$
            \int_{\IR^d} f(x, p_{\sigma^2}*\eta^N_s(x)) \eta^N_s(dx)
            \le
            \bar f(\sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle)
            \sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle .
    $$
    Now, if we let $g(m) = m \bar f(m/ (2 \pi \sigma^2)^{d/2})$,
    which again is finite and nondecreasing,
    then the lemma follows by choosing $K_f = g^{-1}(K_\epsilon)$
    in Lemma~\ref{lem:eta_compact_containment}.
\end{proof}

@@

Similarly,
\begin{align*}
    &
    [M^{N}(f)]_{\tau + t} 
        - [M^{N}(f)]_\tau \big|
    \\ & \qquad =
    \int_{\tau}^{\tau + t}
        \frac{\theta}{N}
        \int_{\IR^d}
        \left\{
            \gamma(x, \eta^N_s)
            \int_{\IR^d} f^2(y) r(y, \eta^N_s) q_\theta(x, dy)
            +
            \mu(x, \eta^N_s) f^2(x)
        \right\}
        \eta^N_s(dx)
    ds .
    \\ & \qquad =
    \int_{\tau}^{\tau + t}
        \frac{\theta}{N}
        \int_{\IR^d}
        \left\{
            \gamma(x, \eta^N_s)
            \left(
                2 f^2(x) r(x, \eta^N_s)
                +
                B_{f^2}(x, \eta^N_s)
            \right)
            +
            f^2(x) F(x, \eta^N_s) / \theta
        \right\}
        \eta^N_s(dx)
    ds ,
\end{align*}
and so using the fact that $\theta/N \to \alpha < \infty$,
an argument similar to 
that for \eqref{eqn:eta_projections_goal1}.
also obtains \eqref{eqn:eta_projections_goal2}.

@@

\begin{remark} \label{remark:BoundedIntegral2}
    Below, we will want to bound
    \[
        \langle \rho_\epsilon * \eta_t, \eta_t \rangle
        =
        \int_{\IR^d} \int_{\IR^d} p_{\epsilon^2}(x, y) \eta_t(dx) \eta_t(dy) .
    \]
    Writing $H(\eta) = \langle \rho_\epsilon * \eta, \eta \rangle$,
    note that since $p_{\epsilon^2}$ is symmetric,
    $H(\eta + \delta_z / N) = H(\eta) + 2 \rho_\epsilon * \eta(z) / N + p_{\epsilon^2}(z, z) / N^2$,
    so following Section~\ref{sec:heuristics},
    \begin{align*}
        \langle \rho_\epsilon * \eta_t, \eta_t \rangle
        &=
        \langle \rho_\epsilon * \eta_0, \eta_0 \rangle
        +
        2 \theta \int_0^t \left \langle \int_{\IR^d}
            \left(
               \rho_\epsilon * \eta_s(y)
               -
               \rho_\epsilon * \eta_s(x)
            \right)
            q_\theta(x, dy)
        , \eta_s(dy) \right \rangle ds
        \\ & \qquad {}
        +
        2 \int_0^t
        \langle
            \rho_\epsilon * \eta_s(x) F(x, \eta_s)
        , \eta_s(dx) \rangle
        ds
        +
        O\left( \frac{1}{\epsilon^d N^2} \right)
        + M_t ,
    \end{align*}
    where $M_t$ is a martingale with quadratic variation
    \begin{align*}
        [M]_t
        &=
        2 \frac{\theta}{N} \int_0^t
        \left \langle
            \int_{\IR^d}
                \rho_\epsilon * \eta_s(y)^2
            q_\theta(x, dy)
        , \eta_s(dx) \right\rangle
        ds
        \\ & \qquad {}
        +
        2 \int_0^t
        \left \langle
            \rho_\epsilon * \eta_s(x)^2 (1 + F(x, \eta_s) / \theta) 
        , \eta_s(dx) \right\rangle
        ds
        +
        O\left( \frac{1}{\epsilon^d N^2} \right) .
    \end{align*}
\end{remark}

>>>>

    First we prove \eqref{eqn:first_moment_rho}.
    Recall that $\smooth{r} \eta(x) = \int p_{\epsilon^2_r}(x - w) \eta(dw)$,
    where $p_t$ is the density of a Gaussian with mean 0 and variance $t$,
    so that applying Fubini, \eqref{eqn:first_moment_rho} is
    \begin{align*}
        \left|
        \int_{\IR^d}
        \theta
            \int_{\IR^d}
                ( p_{\epsilon_r^2}(y - w) - p_{\epsilon_r^2}(x - w) )
            q_\theta(x, dy)
        \eta(dw)
        \right| .
    \end{align*}
    Write $p_{s, x}(\cdot)$ for the density of a Gaussian
    with mean $s \meanq(x)$ and covariance $\epsilon_r^2 I + s \covq(x)$,
    so that $\int p_{\epsilon_r^2}(y - w) q_\theta(x, dy) = p_{1/\theta, x}(w-x)$.
    It therefore suffices to show that for all $x$ and $w \in \IR^d$,
    there exists $K$ such that
    \begin{align} \label{eqn:goal1}
        \left|
        \theta
            \int_{\IR^d}
                ( p_{\epsilon_r^2}(y-w) - p_{\epsilon_r^2}(x-w) )
            q_\theta(x, dy)
        \right|
        &=
        \theta \left|
            p_{1/\theta, x}(w-x) - p_{0, x}(w-x)
        \right|
        \le
        K p_{\epsilon^2_\gamma}(w-x) .
    \end{align}
    However,
    $\theta( p_{1/\theta, x}(z) - p_{0, x}(z) )
    = \partial_s p_{s, x}(z)$ for some $0 \le s \le 1/\theta$.
    Write $\Gamma(s,x) = \epsilon_r^2 I + s \covq(x)$,
    so that
    \begin{align*}
        p_{s, x}(z)
        &=
        \frac{1}{\left(2 \pi |\Gamma(s,x)|\right)^{d/2}}
        \exp\left(
            -\frac{1}{2} (z - s\meanq(x))^T \Gamma(s,x)^{-1} (z - s\meanq(x))
        \right) ,
    \end{align*}
    and note that if $\lambda_i$ are the eigenvalues of $\covq(x)$
    then $|\Gamma(s, x)| = \prod_i (\epsilon_r^2 + s \lambda_i)$,
    and
    $\partial_s |\Gamma(s, x)| = \sum_i \lambda_i |\Gamma(s, x)| / (\epsilon_r^2 + s \lambda_i)$.
    Therefore,
    \begin{align*}
        \partial_s p_{s, x}(z)
        &=
        \left(
            \meanq(x)^T \Gamma(s,x)^{-1} (z - s \meanq)
            +
            (z - s \meanq(x))^T
            \Gamma(s,x)^{-1} \covq(x) \Gamma(s,x)^{-1}
            (z - s \meanq(x))
        \right. \\ & \qquad \qquad \qquad \left. {}
            -
            \sum_i \frac{\lambda_i}{\epsilon_r^2 + s \lambda_i}
        \right)
        p_{s, x}(z) ,
    \end{align*}
    where $z^T$ is the transpose of $z$.
    This implies that
    \begin{align*}
        \frac{
        \theta
            \int_{\IR^d}
                ( p_{\epsilon_r^2}(y-w) - p_{\epsilon_r^2}(x-w) )
            q_\theta(x, dy)
        }{
            p_{\epsilon^2_\gamma}(x-w) .
        }
        =
        h(x-w) e^{k(x-w)},
    \end{align*}
    where $h(z)$ and $k(z)$ are quadratic polynomials in $z$
    whose coefficients depend on $s$ and $x$ but are uniformly bounded for $s \ge 1$
    and $x \in \IR^d$.
    Therefore, \eqref{eqn:goal1} holds for some finite $K$
    if and only if $k(z) < 0$ for all $z$ outside some bounded region.
    This is
    \begin{align*}
        k(z)
        =
        \frac{1}{2\epsilon_\gamma^2} \|z\|^2
        -\frac{1}{2} (z - s\meanq(x))^T \Gamma(s,x)^{-1} (z - s\meanq(x)) ,
    \end{align*}
    which is negative for all $z$ outside a bounded region.
    Since $\inf_z z^T \Gamma(s,x)^{-1} z / \|z\|^2 = 1 / (s \lambda(x) + \epsilon_r^2)$,
    where $\lambda(x) = \sup z^T \covq(x) z / \|z\|^2$ is the largest eigenvalue of $\covq(x)$,
    equation~\eqref{eqn:first_moment_rho} follows from the assumption
    that $\epsilon_r^2 + 2 \sup_x \lambda(x)/\theta < \epsilon_\gamma^2$.
    (Note that we do not yet need the factor of 2.)

@@

\begin{proof}[of Lemma~\ref{lem:tightness_for_F}]
    Let $\bar f(m) = \sup_x \sup_{k \le m} f(x, k)$, which is finite by assumption.
    Since $p_{\sigma^2}*\eta(x) \le \langle 1, \eta \rangle / (2 \pi \sigma^2)^{d/2}$,
    we have that
    $$
        \sup_{0 \le s \le T} \sup_{x \in \IR^d} f(x, p_{\sigma^2}*\eta^N_s(x))
        \le
        \bar f( \sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle  / (2 \pi \sigma^2)^{d/2} ) .
    $$
    This implies that if $0 \le s \le T$ then
    $$
            \int_{\IR^d} f(x, p_{\sigma^2}*\eta^N_s(x)) \eta^N_s(dx)
            \le
            \bar f(\sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle  / (2 \pi \sigma^2)^{d/2} )
            \sup_{0 \le s \le T} \langle 1, \eta^N_s \rangle .
    $$
    Now, if we let $g(m) = m \bar f(m/ (2 \pi \sigma^2)^{d/2})$,
    which again is finite and nondecreasing,
    then the lemma follows by choosing $K_f = g^{-1}(K_\epsilon)$
    in Lemma~\ref{lem:eta_compact_containment}.
\end{proof}

@@

Similarly,
\begin{align*}
    &
    \big| [M^{N}(f)]_{\tau + t} 
        - [M^{N}(f)]_\tau \big|
    \\ & \qquad =
    \int_{\tau}^{\tau + t}
        \frac{\theta}{N}
        \int_{\IR^d}
        \left\{
            \gamma(x, \eta^N_s)
            \int_{\IR^d} f^2(y) r(y, \eta^N_s) q_\theta(x, dy)
            +
            \mu(x, \eta^N_s) f^2(x)
        \right\}
        \eta^N_s(dx)
    ds 
    \\ & \qquad =
    \int_{\tau}^{\tau + t}
        \frac{\theta}{N}
        \int_{\IR^d}
        \left\{
            \gamma(x, \eta^N_s)
            \left(
                2 f^2(x) r(x, \eta^N_s)
                +
                B_{f^2}(x, \eta^N_s)
            \right)
            +
            f^2(x) F(x, \eta^N_s) / \theta
        \right\}
        \eta^N_s(dx)
    ds ,
\end{align*}
and so using the fact that $\theta/N \to \alpha < \infty$,
an argument similar to 
that for \eqref{eqn:eta_projections_goal1}
also obtains \eqref{eqn:eta_projections_goal2}.

@@

\begin{remark} \label{remark:BoundedIntegral2}
    Below, we will want to bound
    \[
        \langle \rho_\epsilon * \eta_t, \eta_t \rangle
        =
        \int_{\IR^d} \int_{\IR^d} p_{\epsilon^2}(x, y) \eta_t(dx) \eta_t(dy) .
    \]
    Writing $H(\eta) = \langle \rho_\epsilon * \eta, \eta \rangle$,
    note that since $p_{\epsilon^2}$ is symmetric,
    $H(\eta + \delta_z / N) = H(\eta) + 2 \rho_\epsilon * \eta(z) / N + p_{\epsilon^2}(z, z) / N^2$,
    so following Section~\ref{sec:heuristics},
    \begin{align*}
        \langle \rho_\epsilon * \eta_t, \eta_t \rangle
        &=
        \langle \rho_\epsilon * \eta_0, \eta_0 \rangle
        +
        2 \theta \int_0^t \left \langle \int_{\IR^d}
            \left(
               \rho_\epsilon * \eta_s(y)
               -
               \rho_\epsilon * \eta_s(x)
            \right)
            q_\theta(x, dy)
        , \eta_s(dy) \right \rangle ds
        \\ & \qquad {}
        +
        2 \int_0^t
        \langle
            \rho_\epsilon * \eta_s(x) F(x, \eta_s)
        , \eta_s(dx) \rangle
        ds
        +
        O\left( \frac{1}{\epsilon^d N^2} \right)
        + M_t ,
    \end{align*}
    where $M_t$ is a martingale with quadratic variation
    \begin{align*}
        [M]_t
        &=
        4 \frac{\theta}{N} \int_0^t
        \left \langle
            \int_{\IR^d}
                \rho_\epsilon * \eta_s(y)^2
            q_\theta(x, dy)
        , \eta_s(dx) \right\rangle
        ds
        \\ & \qquad {}
        +
        4 \int_0^t
        \left \langle
            \rho_\epsilon * \eta_s(x)^2 (1 + F(x, \eta_s) / \theta) 
        , \eta_s(dx) \right\rangle
        ds
        +
        O\left( \frac{1}{\epsilon^d N^2} \right) .
    \end{align*}
\end{remark}
